This script imports a set of georeferenced samples isolated and exported from Echoview for a single transect and allows the user to view the data in 3D to ensure it is free of errors. The script also adjusts the depth of the data based on the transducer offset below the water surface. This is mainly a quality control step. This is done on a transect by transect basis, iterated in a for loop. This script must be run before running the Step4 script, which combines transects into a single output for one survey.

Required input files:

-Exported xyz georeferenced samples isolated from multibeam data for a single transect, from Echoview or similar software, contained within the directory path labeled "path_samples"
-Survey site location with site centroid GPS coordinates, contained within the directory path labeled "path_sites"
-Site bathymetry gridded xyz data for the site in csv format, contained within the directory path labeled "path_bath". This data is generated using the Step1 script.

Created by: Matthew Holland
Email: m.holland@unsw.edu.au
Date: 18 Sept 2020
```{r}
#clear global environment
rm(list = ls())

#user inputs
reef <- "JDN" #specify site name designation in the filename
day <- 8 #specify day of the month the survey took place
month <- 11 #specify month the survey took place
year <- 2019 #specify year the survey took place
dist <- 300 #specify the width of the square survey extent in metres
zone <- 56 #specify UTM grid zone for the site location
hem <- "S" #specify hemisphere the survey was conducted in ("N" or "S")
offset <- 1.5 #depth of transducer below water surface (in m)
path_samples <- "../Data/3D_samples_exports/" #specify the file directory path for the 3D samples
path_sites <- "../Data/Sites.csv" #specify the file directory path for the site coordinates
path_bath <- "../Data/Bathymetry_exports/" #specify the file directory path for the bathymetry xyz csv
```
Load packages and relevant data
```{r}
#check if all required packages are installed. Install them if they are not present. Then load required packages
list.of.packages <- c("sp", "rgl", "knitr", "stringr", "dplyr", "reshape2", "raster")
new.packages <- list.of.packages[!(list.of.packages %in% installed.packages()[,"Package"])]
if(length(new.packages)) install.packages(new.packages)
lapply(list.of.packages, require, character.only = TRUE)

#turn off scientific notation
options(scipen=999)
 
#create a string for folder selection for the date specified
day <- ifelse(nchar(day)<2, paste(0,day,sep=""), day)
month <- ifelse(nchar(month)<2, paste(0,month,sep=""), month)
date <- paste(year,month,day,sep="")

#define CRS for projection
crs_longlat <- paste("+proj=longlat +ellps=WGS84 +south=", ifelse(hem=="S", "T", "F"), " +datum=WGS84", sep="")
crs_utm <- paste("+proj=utm +ellps=WGS84 +south=", ifelse(hem=="S", "T", "F"), " +datum=WGS84", " +zone=", zone, sep="")

#read in reef coordinate data for determining range limits
df.reef <- read.csv(file=path_sites)

#read in bathymetry xyz
bath.files <- list.files(paste(path_bath, sep=""), full.names = T)
bath.files <- subset(bath.files, grepl(as.character(reef), bath.files) & grepl("UTM", bath.files) & grepl("modified", bath.files)) 
df.bath <- read.csv(bath.files[1])

#adjust bathymetry for transducer depth offset
df.bath$z <- df.bath$z - offset

#convert bathymetry to raster
coordinates(df.bath) <- ~ x + y
gridded(df.bath) <- TRUE
df.bath <- raster(df.bath)

#exclude the output files if this transect has already been processed or is not the transect of interest and read in and clean data
numextract <- function(string){str_extract(string, "\\-*\\d+\\.*\\d*")} #function for extracting number from string
ev.files <- list.files(paste(path_samples, date,"/", reef, "/", sep=""), full.names = T) #read in the path for the data file
ev.files <- subset(ev.files, !grepl("processed", ev.files) & grepl(".zip", ev.files))
```
Loop over multiple transects, plot in 3D and output cleaned versions 
```{r}
#for loop to iterate over multiple transects in the same survey
for(i in 1:length(unique(ev.files))){
  
#subset file list to transect of interest  
ev.files.sub <- subset(ev.files, abs(as.numeric(numextract(gsub(".*/|\\.[^.]+$", "", ev.files))))==i) 

#unzip the processed multibeam output file data for the relevant transect
df <- read.table(unzip(ev.files.sub, exdir=paste(path_samples, date,"/", reef, sep="")), sep=",", header=TRUE)
df <- subset(df, Sv > -999)
df <- dplyr::rename(df, Ping_index = 1)

#re-project georeferenced data as UTM
df$Est <- df$Longitude
df$Nrt <- df$Latitude
coordinates(df) <- ~ Est + Nrt
proj4string(df) <- CRS(crs_longlat)
df <- spTransform(df, CRS(crs_utm))

#adjust samples for transducer depth offset
df$Altitude <- df$Altitude - offset
df$Depth <- df$Depth + offset

#exclude all below-bottom samples and within extent of survey area
df$bot.dep <- raster::extract(df.bath, df)*(-1) #extract bottom depth
df <- subset(df, Depth < bot.dep - 1) #subset schools data by bottom depth
df <- as.data.frame(df)

#convert reef centroid data to UTM for subsetting
df.reef$Est <- df.reef$Longitude
df.reef$Nrt <- df.reef$Latitude
coordinates(df.reef) <- ~ Est + Nrt
proj4string(df.reef) <- CRS(crs_longlat)
df.reef <- spTransform(df.reef, CRS(crs_utm))
df.reef <- as.data.frame(df.reef)

#make reef origin for georeferenced points and then subset by spatial extent
df$x <- df$Est - df.reef$Est[df.reef$Code==reef]
df$y <- df$Nrt - df.reef$Nrt[df.reef$Code==reef]
df <- subset(df, abs(x) <= dist/2 & abs(y) <= dist/2) #subset schools data by survey area extent

#remove first ping (necessary sometimes if data is not subset properly by Echoview)
df <- subset(df, Ping_index != min(df$Ping_index))

#Plot in 3D using the RGL package
#generate three-column matrix from DEM
dem <- as.data.frame(df.bath, row.names=NULL, optional=FALSE, xy=TRUE, na.rm=FALSE, long=FALSE)
dem <- data.frame(x=dem$x, y=dem$y, z=dem$z)

#convert to xy coordinates
dem$x <- dem$x - df.reef$Est[df.reef$Code==reef]
dem$y <- dem$y - df.reef$Nrt[df.reef$Code==reef]

#subset to extent of bounding box
dem <- subset(dem, abs(x) < dist/2 & abs(y) < dist/2)

#ensure no NA rows
dem <- dem[complete.cases(dem), ]

#reshape data into an x~y matrix of bathymetry values 
r.dem <- reshape2::acast(dem, x~y, value.var="z")

#calculate reduction factor for z axis to keep coordinates equal
rdctn <- (abs(abs(min(dem$z)) - abs(max(dem$z)))) / (2*dist)

#specify a multiplier for z-axis to accentuate vertical relief (default is 1 for real scale)
mltp <- 2

#create scaled dataframe of samples for plotting
df.plot <- df
df.plot$x <- (df.plot$x + dist/2)/dist
df.plot$y <- (df.plot$y + dist/2)/dist

#generate axis label vectors for 3D plotting
ylabs <- round(seq(from = 1, to = (length(rownames(r.dem))), length=12))
xlabs <- round(seq(from = 1, to = (length(colnames(r.dem))), length=12))

#3D plotting with RGL
knit_hooks$set(webgl = hook_webgl)
nbcol = 100
color=terrain.colors(100)
zcol  = cut(r.dem, nbcol)
persp3d(r.dem, col=color[zcol],clab = "m",shade = 0.1,aspect=c(1,1,rdctn*mltp), axes=FALSE, xlab="x", ylab="y")
axis3d("x", at=seq(0,1, length = length(colnames(r.dem)[xlabs])), labels=round(as.numeric(colnames(r.dem)[xlabs])), box=TRUE, pos=c(0,1,round(min(r.dem))))
axis3d("x", at=seq(0,1, length = length(colnames(r.dem)[xlabs])), labels=round(as.numeric(colnames(r.dem)[xlabs])), box=TRUE, pos=c(0,0,round(min(r.dem))))

axis3d("y", at=seq(0,1, length = length(rownames(r.dem)[ylabs])), labels = round(as.numeric(rownames(r.dem)[ylabs])), box=TRUE, pos=c(0,0,round(min(r.dem))))
axis3d("y", at=seq(0,1, length = length(rownames(r.dem)[ylabs])), labels = round(as.numeric(rownames(r.dem)[ylabs])), box=TRUE, pos=c(1,0,round(min(r.dem))))

axis3d("z", at=round(seq(min(r.dem),0, length = 8)), labels=round(seq(min(r.dem),0, length = 8)), box=TRUE, pos=c(1,0,0))
axis3d("z", at=round(seq(min(r.dem),0, length = 8)), labels=round(seq(min(r.dem),0, length = 8)), box=TRUE, pos=c(0,1,0))
axis3d("z", at=round(seq(min(r.dem),0, length = 8)), labels=round(seq(min(r.dem),0, length = 8)), box=TRUE, pos=c(1,1,0))
axis3d("z", at=round(seq(min(r.dem),0, length = 8)), labels=round(seq(min(r.dem),0, length = 8)), box=TRUE, pos=c(0,0,0))

box3d()

nbcol = 100
color = rev(rainbow(nbcol, start = 0/6, end = 4/6))
zcol  = cut(df.plot$Sv, nbcol)
points3d(x=df.plot$x, y=df.plot$y, z=df.plot$Altitude, col=color[zcol])

#Save the processed acoustic data
write.csv(df, file=paste(path_samples, date,"/", reef, "/", reef, "-", i, "-processed", ".mb.points.csv", sep=""), row.names=F)

}
```

