This script imports a set of bathymetry xyz point cloud data processed from the WASSP DataManager software for multiple surveys carried out at the same site. It rasterizes the seafloor returns into a surface using median values to reduce any impact of anomalously high values (sometimes caused by bubbles under the transducer) and uses ordinary kriging to fill any gaps or holes in the surface. 

It then exports a base bathymetry xyz surface and goes on to import a series of coordinates for artificial objects on the seafloor. It incorporates the 3D outlines of these objects into the bathymetry surface for appropriate subsetting of midwater data.

Required input files:

-A set of xyz exported georeferenced seafloor returns from WASSP DataManager (or similar) contained within the directory path labeled "path_bath"
-Survey site location with site centroid GPS coordinates, contained within the directory path labeled "path_sites"
-A shapefile containing the GPS coordinates of artificial objects, contained within the directory labelled "path_shape"

Created by: Matthew Holland
Email: m.holland@unsw.edu.au
Date: 22 Sept 2020
```{r}
#clear global environment
rm(list = ls())

#user inputs
reef <- "JDN" #specify site name designation in the filename
dist <- 300 #specify the width of the square survey extent in metres
zone <- 56 #specify UTM grid zone for the site location
hem <- "S" #specify hemisphere the survey was conducted in ("N" or "S")
path_sites <- "../Data/Sites.csv" #specify the file directory path for the site coordinates
path_bath <- "../Data/Bathymetry_point_clouds/" #specify the file directory path for the bathymetry xyz csv
path_export <- "../Data/Bathymetry_exports/" #specify the file directory path for the exported school thickess for the survey
path_shape <- "../Data/Reef_shapefiles" #path of the shapefile containing the reef module coordinates
agg.cell <- 5 #specify primary grid cell size for environmental variable detection in m and for the resolution of the aggregated school thickness rasters
res.cell <- 1 #specify resolution of high-resolution raster for rendering 3D outlines of artificial structures etc.
```
Load packages and relevant data
```{r}
#check if all required packages are installed. Install them if they are not present. Then load required packages
list.of.packages <- c("sp", "dplyr", "raster", "ggplot2", "gstat", "rgdal")
new.packages <- list.of.packages[!(list.of.packages %in% installed.packages()[,"Package"])]
if(length(new.packages)) install.packages(new.packages)
lapply(list.of.packages, require, character.only = TRUE)

#define CRS for projection
crs_longlat <- paste("+proj=longlat +ellps=WGS84 +south=", ifelse(hem=="S", "T", "F"), " +datum=WGS84", sep="")
crs_utm <- paste("+proj=utm +ellps=WGS84 +south=", ifelse(hem=="S", "T", "F"), " +datum=WGS84", " +zone=", zone, sep="")
```
Read in and manipulate bathymetry
```{r}
#list the filenames of the relevant bathymetry xyz point clouds
file_list <- list.files(paste(path_bath, reef, "/", sep=""), full.names=T)
file_list <- subset(file_list, !grepl("readme", file_list))

#read in the files and combine into a single dataframe - this step can be slow
df.bath=NULL
for (file in file_list){
      x <- read.csv(file, header=FALSE, sep=",")
      df.bath=rbind(df.bath,x)  
      }

#rename variables appropriately
colnames(df.bath)[1:3] <- c("Long","Lat","z")

#project xyz into UTM
df.bath$Est <- df.bath$Long
df.bath$Nrt <- df.bath$Lat
coordinates(df.bath) <- ~ Est + Nrt
proj4string(df.bath) <- CRS(crs_longlat)
df.bath <- spTransform(df.bath, CRS(crs_utm))

#read in reef coordinate data for determining range limits
df.reef <- read.csv(file=path_sites)

#convert reef centroid data to UTM for subsetting
df.reef$Est <- df.reef$Longitude
df.reef$Nrt <- df.reef$Latitude
coordinates(df.reef) <- ~ Est + Nrt
proj4string(df.reef) <- CRS(crs_longlat)
df.reef <- spTransform(df.reef, CRS(crs_utm))
df.reef <- as.data.frame(df.reef)

#make reef origin for points
df.bath$x <- df.bath$Est - df.reef$Est[df.reef$Code==reef]
df.bath$y <- df.bath$Nrt - df.reef$Nrt[df.reef$Code==reef]

#subset bathymetry xyz data to survey extent box
buf <- 10 #define the number of buffer cells around each side of the survey extent
df.bath <- subset(df.bath, abs(x) <= (dist/2)+buf*agg.cell & abs(y) <= dist/2++buf*agg.cell)

#define raster
r.base <- raster(ncols=dist/agg.cell+buf*2, nrows=dist/agg.cell+buf*2, xmn=(-1)*dist/2-agg.cell*buf, xmx=dist/2+agg.cell*buf, ymn=(-1)*dist/2-agg.cell*buf, ymx=dist/2+agg.cell*buf)

#rasterise bathymetry
df.bath <- as.data.frame(df.bath)
coordinates(df.bath) <- ~ x + y
r.bath <- rasterize(df.bath, r.base, field = df.bath$z, fun=function(x,...) {median(x, na.rm=T)}, update = TRUE)

#remove any above 0 errors in the data
r.bath[r.bath >= 0] <- NA

#smooth the rasterised data with a focal function to reduce any anomalous values
r.bath <- focal(r.bath, w=matrix(1, 3, 3), function(x) median(x, na.rm=TRUE))

#plot the bathymetry raster 
plot(r.bath)

#convert to dataframe
sub.rast <- as.data.frame(r.bath, row.names=NULL, optional=FALSE, xy=TRUE, na.rm=FALSE, long=FALSE)
sub.rast <- dplyr::rename(sub.rast, z = layer)

#plot the results
ggplot(sub.rast, aes(x,y,fill=z))+
  geom_tile()+
  coord_equal()+
  theme_bw()+
  scale_x_continuous(expand=c(0,0))+
  scale_y_continuous(expand=c(0,0))+
  scale_fill_gradientn(colours = terrain.colors(10))
```
Use ordinary kriging to interpolate over any missing data (holes) in bathymetry raster
```{r}
krig.bath <- subset(sub.rast, !(is.na(sub.rast$z)))
krig.bath <- krig.bath[sample(nrow(krig.bath), nrow(krig.bath/2)), ]

#define coordinates
coordinates(krig.bath) <- ~ x + y

#create variogram model
lzn.vgm <- variogram(z~1, data=krig.bath, cutoff=100) # calculates sample variogram values 
lzn.fit <- fit.variogram(lzn.vgm, vgm(c( "Exp", "Sph", "Gau", "Mat"))) # fit model

#plot variogram model
plot(lzn.vgm, lzn.fit) # plot the sample values, along with the fit model

#create empty raster dataframe
krig.grid <- as.data.frame(r.base, row.names=NULL, optional=FALSE, xy=TRUE, na.rm=FALSE, long=FALSE)

#define coordinates
coordinates(krig.grid) <- ~ x + y

#krig into empty raster dataframe
lzn.kriged <- as.data.frame(krige(z ~ 1, krig.bath, krig.grid, model=lzn.fit, maxdist=20))

#matching kriged pixels to original bathy data
krig.cells <- subset(sub.rast, (is.na(sub.rast$z)))
sub.rast <- subset(sub.rast, !(is.na(sub.rast$z)))

#unique matching variables
krig.cells <- within(krig.cells, Uniq.ID <- paste(x, y, sep=','))
lzn.kriged <- within(lzn.kriged, Uniq.ID <- paste(x, y, sep=','))
sub.rast <- within(sub.rast, Uniq.ID <- paste(x, y, sep=','))

#match NA cells with kriged data
krig.cells$z <- lzn.kriged$var1.pred[match(krig.cells$Uniq.ID, lzn.kriged$Uniq.ID)]

#bind dataframes to replace missing data
sub.rast <- rbind(sub.rast, krig.cells)

#plot the results of filling the holes in the data
ggplot(sub.rast, aes(x,y,fill=z))+
  geom_tile()+
  coord_equal()+
  theme_bw()+
  scale_x_continuous(expand=c(0,0))+
  scale_y_continuous(expand=c(0,0))+
  scale_fill_gradientn(colours = terrain.colors(10))
```
Create export file for base bathymetry XYZ surface and save file
```{r}
#create new dataframe
df.export <- sub.rast

#convert coordinate system back to UTM
df.export$x <- df.export$x + df.reef$Est[df.reef$Code==reef]
df.export$y <- df.export$y + df.reef$Nrt[df.reef$Code==reef]

#keep only three columns (xyz)
df.export <- dplyr::select(df.export, c("x", "y", "z"))

#Save .csv file of unmodified bathymetry (no added 3D structures) for use in the Step4 script
write.csv(df.export, paste(path_export, reef, "-base", ".csv", sep=""), row.names = F)
```
Create modified bathymetry surface containing 3D shapes of structures for subsetting in Echoview. This step may not always be necessary. 
```{r}
#replace bathymetry raster with kriged version (holes filled)
coordinates(df.export) <- ~ x + y
gridded(df.export) <- TRUE
r.bath <- raster(df.export)
df.export <- as.data.frame(df.export)

#disaggregate raster so that structure outlines can be resolved in higher resolution 
r.bath.disagg <- disaggregate(r.bath, fact=agg.cell/res.cell, method="bilinear")

#enter reef module dimensions
wide <- 4 #module width (assuming square)
high <- 5 #module height
bot.dep <- -30 #bottom depth
top.ind <- 1.25 #top indentation
top.wid <- 2 #top width
ant <- 3.5 #antenna height
ant.wid <- 0.25 #antenna width
res <- 0.25 #resolution of raster
cnr <- (wide-top.ind)/2 #corner indentation

#read shapefile of reef module coordinates
df.outlines <- as.data.frame(readOGR(dsn=path_shape,layer="Actual_Modules_GDA94_Z56"))
df.outlines <- subset(df.outlines, Set == "North Set")

#calculate boundaries of the reef modules
top <- df.outlines$Pos_Northi + 0.5*wide
bottom <- df.outlines$Pos_Northi - 0.5*wide
left <- df.outlines$Pos_Eastin - 0.5*wide
right <- df.outlines$Pos_Eastin + 0.5*wide

#combine these outlines into a dataframe
df.outlines <- data.frame(top=top, bottom=bottom, left=left, right=right, Est.mid=df.outlines$Pos_Eastin , Nrt.mid=df.outlines$Pos_Northi, tow=df.outlines$Tower)

#create empty base raster to fill with new updated bathymetry, including modules
grid <- as.data.frame(r.bath.disagg, row.names=NULL, optional=FALSE, xy=TRUE, na.rm=FALSE, long=FALSE)
grid <- dplyr::rename(grid, Est=x, Nrt=y)

#create initial shape of reefs from shape bounds
for(i in 1:nrow(df.outlines)){
  grid$z <- ifelse(grid$Est >= df.outlines$left[i] & grid$Est <= df.outlines$right[i] &
                     grid$Nrt >= df.outlines$bottom[i] & grid$Nrt <= df.outlines$top[i],
                   grid$z + high, 
                   grid$z)
}


#create towers
for(i in 1:nrow(df.outlines)){
  grid$z <- ifelse(abs(grid$Est - df.outlines$Est.mid[i]) < ant.wid & abs(grid$Nrt - df.outlines$Nrt.mid[i]) < ant.wid & df.outlines$tow[i] == "Yes",
                   grid$z + ant + top.ind, 
                   grid$z)
}
```
Make DEM
```{r}
dem <- data.frame(x=grid$Est, y=grid$Nrt, z=grid$z)
r.dem <- reshape2::acast(dem, x~y, value.var="z")

#4D plotting function
levelpersp <- function(x, y, z, colors=terrain.colors, ...) {
  ## getting the value of the midpoint
  zz <- (z[-1,-1] + z[-1,-ncol(z)] + z[-nrow(z),-1] + z[-nrow(z),-ncol(z)])/4
  ## calculating the breaks
  breaks <- hist(zz, plot=FALSE)$breaks
  ## cutting up zz
  cols <- colors(length(breaks)-1)
  zzz <- cut(zz, breaks=breaks, labels=cols)
  ## plotting
  persp(x, y, z, col=as.character(zzz), ...)
  ## return breaks and colors for the legend
  list(breaks=breaks, colors=cols)
}

#calculate reduction factor for z axis to keep coordinates equal
rdctn <- (abs(abs(min(dem$z, na.rm=T)) - abs(max(dem$z, na.rm=T)))) / (2*dist)

#specift a multiplier for z-axis (default is 1 for real scale)
mltp <- 2

#3D plot for visualising the result
levelpersp(sort(unique(grid$Est)), sort(unique(grid$Nrt)), r.dem,  border=NA, phi=25, theta=135, shade=0.15, expand=rdctn*mltp, axes=TRUE, ticktype="detailed", nticks=2)
```
Save updated bathymetry for subsetting in Echoview and R
```{r}
#rename variables for xyz format
df.export2 <- dplyr::rename(grid, x = Est, y = Nrt)

#Save .csv file as UTM
write.csv(df.export2, paste(path_export, reef, "-modified", "-UTM", ".csv", sep=""), row.names = F)

#Save .csv file as longlat for use as a boundary in Echoview
coordinates(df.export2) <- ~ x + y
proj4string(df.export2) <- CRS(crs_utm)
gridded(df.export2) <- TRUE
df.export2 <- raster(df.export2)
df.export2 <- projectRaster(df.export2, crs=CRS(crs_longlat))
df.export2 <- as.data.frame(df.export2, row.names=NULL, optional=FALSE, xy=TRUE, na.rm=FALSE, long=FALSE)
df.export2 <- data.frame(Latitude=df.export2$y, Longitude=df.export2$x, Depth=abs(df.export2$z))
write.csv(df.export2, paste(path_export, reef, "-modified", "-longlat", ".csv", sep=""), row.names = F)
```

```


