This script imports a set of cleaned transect samples isolated and exported from Echoview and processed in R with the Step2 script. It processes the 3D xyz georeferenced target data for a single survey and writes it into a dataframe on a transect by transect basis. 

Required input files:

-A full set of 3D xyz exported georeferenced samples isolated from multibeam data for a single survey, from Step3 script, contained within the directory path labeled "path_samples"
-Survey site location with site centroid GPS coordinates, contained within the directory path labeled "path_sites"
-Site bathymetry gridded xyz data for the site in csv format, from Step1 script, contained within the directory path labeled "path_bath"
-Cruise track GPS data processed by Step2 script, contained within the directory path labeled "path_tracks"

Created by: Matthew Holland
Email: m.holland@unsw.edu.au
Date: 21 Sept 2020
```{r}
#clear global environment
rm(list = ls())

#user inputs
reef <- "JDN" #specify site name designation in the filename
day <- 8 #specify day of the month the survey took place
month <- 11 #specify month the survey took place
year <- 2019 #specify year the survey took place
dist <- 300 #specify the width of the square survey extent in metres
zone <- 56 #specify UTM grid zone for the site location
hem <- "S" #specify hemisphere the survey was conducted in ("N" or "S")
offset <- 1.5 #depth of transducer below water surface (in m)
path_samples <- "../Data/3D_samples_exports/" #specify the file directory path for the GPS waypoints
path_sites <- "../Data/Sites.csv" #specify the file directory path for the site coordinates
path_bath <- "../Data/Bathymetry_exports/" #specify the file directory path for the bathymetry xyz csv
path_tracks <- "../Data/Track_exports/" #specify the file directory path for the cruise track GPS waypoints
path_export <- "../Data/School_thickness_exports/" #specify the file directory path for the exported school thickess for the survey
sch.cell <- 1 #specify the resolution at which school thickness will be initially rasterised
agg.cell <- 5 #specify primary grid cell size for environmental variable detection in m and for the resolution of the aggregated school thickness rasters
mindep <- 10 #specify minimum depth below transducer that all shallower targets will be cropped to
swath <- 120 #specify multibeam echosounder swath width in degrees
```
Load packages and relevant data
```{r}
#check if all required packages are installed. Install them if they are not present. Then load required packages
list.of.packages <- c("sp", "ggplot2", "raster", "dplyr", "reshape2", "pracma", "rgeos", "maptools", "stringr", "rgl", "tidyr")
new.packages <- list.of.packages[!(list.of.packages %in% installed.packages()[,"Package"])]
if(length(new.packages)) install.packages(new.packages)
lapply(list.of.packages, require, character.only = TRUE)

#turn off scientific notation
options(scipen=999)

#create a string for folder selection for the date specified
day <- ifelse(nchar(day)<2, paste(0,day,sep=""), day)
month <- ifelse(nchar(month)<2, paste(0,month,sep=""), month)
date <- paste(year,month,day,sep="")

#define CRS for projection
crs_longlat <- paste("+proj=longlat +ellps=WGS84 +south=", ifelse(hem=="S", "T", "F"), " +datum=WGS84", sep="")
crs_utm <- paste("+proj=utm +ellps=WGS84 +south=", ifelse(hem=="S", "T", "F"), " +datum=WGS84", " +zone=", zone, sep="")

#create a list of the relevant 3D samples files to load
file_list <- list.files(paste(path_samples, date,"/", reef, "/", sep=""), full.names = T) #read in the path for the data file
file_list <- subset(file_list, grepl("processed", file_list)) #exclude unprocessed data

#load in the relevant pre-processed 3D samples from Step2 for a single survey - this step can be slow
numextract <- function(string){str_extract(string, "\\-*\\d+\\.*\\d*")} #function for extracting number from string
df=NULL
for (file in file_list){
  x <- read.csv(paste(file), header=TRUE, sep=",")
  x$trans.num <- abs(as.numeric(numextract(gsub(".*/|\\.[^.]+$", "", file))))
  df=rbind(df,x)
  rm(x)
}

#subset schools by bounding box of area
df <- subset(df, abs(x) <= dist/2 & abs(y) <= dist/2)
```
Read in reef site info in order to inform subsetting of boundary area and generate orthogonal coordinates
```{r}
#read in reef coordinate data for determining range limits
df.reef <- read.csv(file=path_sites)

#convert reef centroid data to UTM for subsetting
df.reef$Est <- df.reef$Longitude
df.reef$Nrt <- df.reef$Latitude
coordinates(df.reef) <- ~ Est + Nrt
proj4string(df.reef) <- CRS(crs_longlat)
df.reef <- spTransform(df.reef, CRS(crs_utm))
df.reef <- as.data.frame(df.reef)
```
Read in and manipulate bathymetry xyz data
```{r}
bath.files <- list.files(paste(path_bath, sep=""), full.names = T)
bath.files <- subset(bath.files, grepl(as.character(reef), bath.files) & grepl("base", bath.files)) 
df.bath <- read.csv(bath.files[1])
df.bath <- dplyr::rename(df.bath, Est = x, Nrt = y)

#include transducer depth offset in bathymetry
df.bath$z <- df.bath$z - offset

#make reef origin for points
df.bath$x <- df.bath$Est - df.reef$Est[df.reef$Code == reef]
df.bath$y <- df.bath$Nrt - df.reef$Nrt[df.reef$Code == reef]

#define coordinates
coordinates(df.bath) <- ~ x + y

#define raster
buf <- 1 #define the number of buffer cells around each side of the survey extent
r.base <- raster(ncols=dist/agg.cell+buf*2, nrows=dist/agg.cell+buf*2, xmn=(-1)*dist/2-agg.cell*buf, xmx=dist/2+agg.cell*buf, ymn=(-1)*dist/2-agg.cell*buf, ymx=dist/2+agg.cell*buf)

#rasterise bathymetry
r.bath <- rasterize(df.bath, r.base, field = df.bath$z, update = TRUE)

#convert raster bathymetry to a dataframe and overwrite df.bath
df.bath <- as.data.frame(r.bath, row.names=NULL, optional=FALSE, xy=TRUE, na.rm=FALSE, long=FALSE)
df.bath <- dplyr::rename(df.bath, z = layer)

#plot the bathymetry
ggplot(df.bath, aes(x,y,fill=z))+
  geom_raster()+
  coord_equal()+
  theme_bw()+
  scale_x_continuous(expand=c(0,0))+
  scale_y_continuous(expand=c(0,0))+
  scale_fill_gradientn(colours = terrain.colors(10))
```
3D plotting for visual inspection of all the data from a single survey
```{r}
dem <- df.bath #make matrix from DEM
dem <- subset(dem, abs(x) <= dist/2 & abs(y) <= dist/2) #subset to bounding box extent
dem <- dem[complete.cases(dem), ] #ensure no NA rows
r.dem <- reshape2::acast(dem, x~y, value.var="z") #reshape dataframe to matrix

#calculate reduction factor for z axis to keep coordinates equal
rdctn <- (abs(abs(min(dem$z)) - abs(max(dem$z)))) / (2*dist)

#specify a multiplier for z-axis (default is 1 for real scale)
mltp <- 2

#create scaled dataframe of samples for plotting
df.plot <- df
df.plot$x <- (df.plot$x + dist/2)/dist
df.plot$y <- (df.plot$y + dist/2)/dist

#generate axis label vectors for 3D plotting
ylabs <- round(seq(from = 1, to = (length(rownames(r.dem))), length=12))
xlabs <- round(seq(from = 1, to = (length(colnames(r.dem))), length=12))

#3D plotting
knit_hooks$set(webgl = hook_webgl)
nbcol = 100
#color = rev(rainbow(nbcol, start = 0/6, end = 4/6))
color=terrain.colors(100)
zcol  = cut(r.dem, nbcol)
persp3d(r.dem, col=color[zcol],clab = "m",shade = 0.1,aspect=c(1,1,rdctn*mltp), axes=FALSE, xlab="x", ylab="y")
axis3d("x", at=seq(0,1, length = length(colnames(r.dem)[xlabs])), labels=colnames(r.dem)[xlabs], box=TRUE)
axis3d("y", at=seq(0,1, length = length(rownames(r.dem)[ylabs])), labels = rownames(r.dem)[ylabs], box=TRUE)
axis3d("z", at=round(seq(min(r.dem),0, length = 8)), labels=round(seq(min(r.dem),0, length = 8)), box=TRUE)
box3d()

#plot points (schools)
nbcol = 100
color = rev(rainbow(nbcol, start = 0/6, end = 4/6))
zcol  = cut(df.plot$Sv, nbcol)
points3d(x=df.plot$x, y=df.plot$y, z=df.plot$Altitude, col=color[zcol],aspect=c(1,1,mltp))
```
Prepare 3D samples dataframe and base raster for spatial subsetting and rasterization of school distribution data
```{r}
#subset by minimum depth
df <- subset(df, Depth > mindep)

#define coordinates for schools
coordinates(df) <- ~ x + y

#extract bottom depth
df$bot.dep <- raster::extract(r.bath, df)*(-1)

#subset schools data by bottom depth
df <- subset(df, Depth < bot.dep - 1)
```
Load in track files and buffer around them based on depth and trigonometry on a per point basis to reveal surveyed versus unsurveyed sections
```{r}
#read list of filenames for processed cruise track files
track.files <- list.files(paste(path_tracks, date,"/", reef, "/", sep=""), full.names = T)
track.files <- subset(track.files, grepl("processed", track.files)) #exclude the input files if this track has not been processed

#sequentially read in cruise track files for the specified survey
df.tracks=NULL
for (track in track.files){
  x <- read.csv(track, header=TRUE, sep=",")
  df.tracks=rbind(df.tracks,x)  
}

#plot he tracks as they have been read in
ggplot(df.tracks, aes(x, y, colour=as.factor(track)))+
  geom_point()+
  coord_equal()+
  theme_bw()

#extract bottom depth at each point along the cruise track
coordinates(df.tracks) <- ~ x + y
df.tracks$depth <- raster::extract(r.bath, df.tracks)

#calculate mean depth of the track for working out swath width with trig and 120 degree beam
df.tracks$buff.wid <- abs(mindep)*tand(120/2)

#reduce the size of the dataframe by extracting every 50th row
df.tracks <- df.tracks[seq(1, nrow(df.tracks), 50), ]

#generate polygons from buffering around the cruise track
spList = vector("list", length(df.tracks))
for (i in 1:length(df.tracks)) {
  a <- rgeos::gBuffer(df.tracks[i,], width = (df.tracks$buff.wid[i]))
  a$id = df.tracks$track[i]
  spList[[i]] <- a
}
spListAll <- do.call("rbind", spList)

#merge polygons within transects and plot for visual inspection
spListAll <- maptools::unionSpatialPolygons(spListAll, spListAll$id)
plot(spListAll)

#create spatial p[olygons data frame
df.buff <- SpatialPolygonsDataFrame(Sr=spListAll, data=data.frame(id=1:length(unique(spListAll@polygons))),FALSE)

```
Rasterise 3D school thickness data initially, then min, then max depth
```{r}
#create base raster grid
r.base <- raster(ncols = dist/sch.cell, nrows = dist/sch.cell, xmn=(-1)*dist/2, xmx=dist/2, ymn=(-1)*dist/2, ymx=dist/2)

#bin depth by 1 m bins
df$depth.bin <- cut(df$Depth, breaks=seq(0,1000,1), labels=seq(0.5, 999.5, 1))

#list unique transects
sch.num <- unique(df$trans.num)

#define empty raster stacks
vol.rast <- stack()
min.rast <- stack()
max.rast <- stack()

#rasterise schools on a transect by transect basis - this step can be slow
for (sch in sch.num){
  
  #subset 3D data by transect
  sch.subset <- subset(df, trans.num == sch)
  
  #rasterize using the sum method
  a <- rasterize(sch.subset, r.base, field = sch.subset$depth.bin, fun=function(x,...) {length(unique(x))}) 
  b <- rasterize(sch.subset, r.base, field = sch.subset$Depth, fun=function(x,...) {min(x, na.rm=T)}) 
  c <- rasterize(sch.subset, r.base, field = sch.subset$Depth, fun=function(x,...) {max(x, na.rm=T)}) 
  
  #subset buffer SPDF to correct buffer and rasterize buffer
  buff.subset <- subset(df.buff, id == sch)
  buff.subset <- rasterize(buff.subset, r.base)

  #assign 0 to NA cells in school thickness transect
  a[is.na(a)] <- 0
  
  #then crop transect by buffer
  a[is.na(buff.subset)] <- NA
  b[is.na(buff.subset)] <- NA
  c[is.na(buff.subset)] <- NA

  #combine rasters into raster stacks
  vol.rast <- stack(vol.rast, a)
  min.rast <- stack(min.rast, b)
  max.rast <- stack(max.rast, c)
   
}

#rename rasters in stacks to correspond with transects
names(vol.rast) <- as.numeric(unique(df$trans.num))
names(min.rast) <- as.numeric(unique(df$trans.num))
names(max.rast) <- as.numeric(unique(df$trans.num))

#plot the raster stacks
plot(vol.rast)
plot(min.rast)
plot(max.rast)

#create dataframes from the stacks
vol.rast.2D <- as.data.frame(vol.rast, row.names=NULL, optional=FALSE, xy=TRUE, na.rm=FALSE, long=FALSE)
min.rast.2D <- as.data.frame(min.rast, row.names=NULL, optional=FALSE, xy=TRUE, na.rm=FALSE, long=FALSE)
max.rast.2D <- as.data.frame(max.rast, row.names=NULL, optional=FALSE, xy=TRUE, na.rm=FALSE, long=FALSE)

#save transect by transect version for raster analysis
vol.rast.2D <- tidyr::gather(vol.rast.2D, trans, vol, 3:ncol(vol.rast.2D))
min.rast.2D <- tidyr::gather(min.rast.2D, trans, min, 3:ncol(min.rast.2D))
max.rast.2D <- tidyr::gather(max.rast.2D, trans, max, 3:ncol(max.rast.2D))

#create final combined dataframe
rast.2D <- cbind(vol.rast.2D, min=min.rast.2D$min, max=max.rast.2D$max)
rm(vol.rast.2D, min.rast.2D, max.rast.2D)

#rename transects (remove the X associated with the raster stack names)
rast.2D$trans <- gsub("X", "", rast.2D$trans)

#add bathymetry to the dataframe as a variable
coordinates(rast.2D) <- ~ x + y
rast.2D$bath <- raster::extract(r.bath, rast.2D)*(-1)
rast.2D <- as.data.frame(rast.2D)
```
Plot the three rasterized variables
```{r}
#define colour palette
jetcols <- c("#00007F", "blue", "#007FFF", "cyan", "#7FFF7F", "yellow", "#FF7F00", "red", "#7F0000")

#plot volume (thickness)
ggplot()+
geom_tile(data=subset(rast.2D,!(is.na(vol))), aes(x=x, y=y, fill=vol))+
coord_equal()+
scale_fill_gradientn(colours=jetcols)+
scale_x_continuous(expand=c(0,0))+
scale_y_continuous(expand=c(0,0))+
theme_bw()

#plot min depth
ggplot()+
geom_tile(data=subset(rast.2D,!(is.na(vol))), aes(x=x, y=y, fill=min))+
coord_equal()+
scale_fill_gradientn(colours=jetcols)+
scale_x_continuous(expand=c(0,0))+
scale_y_continuous(expand=c(0,0))+
theme_bw()

#plot max depth
ggplot()+
geom_tile(data=subset(rast.2D,!(is.na(vol))), aes(x=x, y=y, fill=max))+
coord_equal()+
scale_fill_gradientn(colours=jetcols)+
scale_x_continuous(expand=c(0,0))+
scale_y_continuous(expand=c(0,0))+
theme_bw()
```
Prepare transect dataframe to be saved so that it can be used in another script
```{r}
#add in date and site as variables in dataframe
rast.2D$date <- paste(day, month, year, sep="-")
rast.2D$site <- reef

#remove all NA (unsurveyed) cells to save on file size
rast.2D <- subset(rast.2D, !(is.na(rast.2D$vol)))

#save data as .csv file
Filename <- "trans-grid-df"
write.csv(rast.2D, file=paste(path_export, Filename, "-", reef, "-", rast.2D$date[1], ".csv",sep=""), row.names=F)
```